The role of standing variation in parallel adaptation

### 2014-09-22 standing-variation-paper.tex
changes throughout, added G6PD_chartimes_sd_sb.pdf

@@

\plr{the following needs better explanation, maybe a figure?}
Before $t=0$, on the other hand, the mutation is deleterious.
We have assumed that $s_d$ is not too small
so that the genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well--approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Poisson Mapping Theorem, a Poisson process with rate we define to be $\lambda_0$.
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}
Note that this is closely related to the equilibrium frequency of a
allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}

\plr{XXX More discussion on whether this is a good approximation?  
Compute typical size and radius of a standing cluster?}

\paragraph{Geographic spread of alleles} 
Once an allele has become locally established it can begin to spread across space. 
We assume that the allele, once established, 
quickly settles down to spread spatially as a traveling wave of constant speed. 
The behavior of this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. 
Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
See \citet{ralph2010parallel} for a fuller review of these travelling waves.
Note that the speed of the wave will vary
with details of the space that individuals migrate across
\citep[e.g.\ see ][ for comparisons to migration on discrete grids]{Slatkin-speed:76,SlatkinCharlesworth:78}.
% However, the velocity will usually be primarily determined by the product of $\sigma^2$ and $s$.

@@

% In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
% speeds, as can arise due to fat-tailed dispersal kernels. 
% Here we restict ourselves to constant wave speeds for simplicity but similar
% calculations could be done by expansion of the results of the previous paper.

@@

Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutatoin at different locations within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Three G6PD deficiency alleles that also confer resistance to malaria 
are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years  \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
\plr{what are Ruwende's results? feels a bit vague here} \\

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. 
The pie chart shows the frequency of G6pd-deficiency alleles. 
The size of the pie chart indicates the number of G6pd-deficient individuals sampled.
Countries with endemic malaria are colored yellow. 
Figure taken from \citet{Howes-g6pd-variants}
\url{http://www.malariajournal.com/content/12/1/418}. 
\gc{Not sure whether to include this.}
\plr{I like it.}
} \label{fig-G6PD-map}
\end{center}
\end{figure}

@@

Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
As it spreads, it will cover a distance $L$ in time $L/v$.
The number of other mutations appearing in the circle it has covered up until this time is Poisson with mean
$\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
(and $\lambda_0 2 L + \lambda \pi L^2 /v$ in one dimension).
% $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
\[
    \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
\]
then $\chi$ gives the distance spread unobstructed by the descendants of a new mutant
before it is expected that one other successful mutation would have arisen in the area covered so far.
The explicit formula for $\chi$ is cumbersome; here we omit it.
(In one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
we can rewrite this as
\begin{equation} \label{eqn:defines_chi}
%    \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^2 + \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^3 
%        = \frac{ - \log(1-s_d)^3 }{ 4 s_b^2 \sigma^2 \pi \rho \mu } .
   \frac{\sqrt{2s_b} }{-\log(1-s_d) } \chi^2 + \frac{1}{\sigma} \chi^3 = \frac{1}{\sqrt{2s_b} \rho\mu\pi}
\end{equation}
From this we see that $\chi$ decreases with $\rho$ and $\mu$.
Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
$\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma)$.
On the other hand, if the mutant allele is highly deleterious before $t=0$,
then the characteristic length is close to the value from \citet{ralphcoop2010}:
$\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d))$.

XXX could do calculations as in "panmixia" here, e.g.\ if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
$f = 1-\gamma f^{3/2}$, where
$\gamma = (4 \pi^{5/2} / \sqrt{2}) \mu^2 s_b^{3/2} \rho^2 / \sigma \log(1/(1-s_d))$.
Then if $\gamma \ll 1$, $f \approx 1$, while if $\gamma \gg 1$, then $f \approx 0$.

By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
and occur within the first $\chi/v$ generations.
Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
we expect to find one mutational origin;
we denote the probability that this mutation is from standing variation by $\pi_0$.
Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
and it is an property of competing Poisson processes that
\begin{align} \label{eqn:pizero}
    \pi_0 = \lambda_0 \pi \chi^2 .
\end{align}
(In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)

In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths as a function 
of the G6pd parameters.

@@

The quantity that is perhaps easiest to compute is the mean time until adaptation.
If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as can be seen from Figure~\ref{fig:cartoon},
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.
In Figure \ref{G6PD_chartimes} we show the mean time until adaptation for various values of the parameters.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_chartimes}   %{charlen-by-sd-limit}
\caption{ %
{\bf Mean adaptation times,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
} \label{G6PD_chartimes}
\end{center}
\end{figure}

@@

We have defined $\lambda_0$ to be the mean density of standing variants that reach local fixation.
Define $\nu_+$ to be the mean density of new mutants whose offspring fix locally.
Since the probability that a mutant arising at $(x,t)$ is lucky enough to be born in a location not already occupied by mutants
is $\P\{ \tau > t \}$,
we can see  $\nu_+ = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
$\nu_+ = \lambda \E[\tau] $.
Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
\begin{equation}
    \nu_+ = (\mbox{mean density of new patches}) = \frac{1}{1-\log(1-s_d) \E[\tau]} .
\end{equation}
Since the mean proportion of patches that come from standing variation is $\lambda_0 / (\lambda_0 + \nu_+)$,
this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
There are $\lambda_0 + \nu_+$ patches per unit area, so
the typical patch (i.e.\ with distribution given by the Palm measure) occupies area $1/(\lambda_0 + \nu_+)$.

We can also find the mean proportion of space covered by standing variants.
If by time $t$ a geographic location has not yet been reached by the mutation,
the probability it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$, 
and the probability it is reached by a new variant is $\lambda \pi v^2 t^2 dt$.
Therefore, as is standard for competing exponentials,
the probability a given location is reached first by a standing variant,
and therefore the mean proportion of space covered  by standing variants,
is
\begin{equation}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt .
\end{equation}
\marginnote{put in this value in terms of Gaussian CDF?}
To evaluate this integral, again see Appendix \ref{apx:integrals}.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
\caption{ %
{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the G6PD example.  XXX say more.
} \label{G6PD_standing_var_proportion}
\end{center}
\end{figure}

{\tt Simulations:} 
compare mean density against $s_d$ at different values of $\sigma$; 
omit variability if needed for readability. 
Compare to panmictic results.

\subsubsection{Multiple variant types}

One interesting problem that we can seek to address with this work is
the extent to which the degree of pleiotropy biases 
adaptation towards the repeated use of particular subset of loci
(i.e.\ convergence at genetic level). 
While many alleles may confer the benefical phenotype, 
not all will contribute to adaptation if they have negative pleiotropic consequences.
There are at least two ways that negative pleiotropy can contribute to high rates of convergence 
when adapting if a single change is sufficient for adaptation. 
First, negative pleiotropic effects can reduce 
the overall beneficial selection coefficient of an allele in the new environment,
making them unlikely to become established and slow to spread 
(and in the worst case making them deleterious). 
This effect has been studied by a number of authors \citep{}. 
A second contribution is that alleles that have
less negative pleiotropy are more likely to be present as standing variation 
before the environmental shift.
This effect seems less well studied, 
perhaps because many modeling approaches use the strong selection, weak mutation limit 
and so ignore standing variation. 
\gc{Is this fair?}
\plr{Not sure what papers you are thinking of.}

Here we focus primarily on the second effect. 
Let's imagine for the moment that all classes of beneficial allele 
have the same beneficial selection coefficient. 
Each class of mutations could have its own mutation rate $\mu_j$ and
have a disadvantage $s_{d,j}$ prior to the switch. 
As they have the same beneficial selection coefficient after the switch, 
all of the waves travel outward at a rate $v$. 
Then the density of type $k$ standing variants per unit area 
and the input rate of \textit{de novo} variants per unit area per generation are, respectively,
\begin{align}
  \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } \qquad \text{and} \qquad    \lambda_{j} = \mu_j \rho s_b .
\end{align}
Using these rates it follows that, at the time when every location has been reached by an adaptive allele,
the proportion of the species range covered by alleles of type $k$ is
\begin{equation} \label{eqn-prop-space-allele-k}
    p_k = \int_0^\infty  \left( 2 \lambda_{0,k} \pi v^2 t + \lambda_k \pi v^2 t^3 \right)  
    \; \exp \left( - \sum_j ( \lambda_{0,j} \pi v^2 t^2 + \lambda_j \pi v^2 t^3 / 3 ) \right) dt .
\end{equation}
If we only allow standing variation this collapses to 
\begin{equation}  \label{eqn-prop-space-allele-k-standing-room-only}
    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) } ,
\end{equation}
while if we only allow new variation, 
i.e.\ if all variants are highly deleterious before the environment switches, 
then $p_k =\mu_j / (\sum_k  \mu_k)$.

>>>>

We show that convergent adaptation within wide-spread species is likely to be
common, as has already been seen for a number of traits. 
On the basis of this we argue, in our discussion, that the genetics of
adaptation may often be
geographically local, and that we should expect to see relatively few
wide-spread sweeps even when selection pressures are wide-spread. 
We discuss some implications, and history, of this view for the evolutionary
coherence of species and molecular evolution.

@@

Before $t=0$, on the other hand, the allele is deleterious.
The genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for some time (note that we have assumed that $s_d$ is not too small).
Since the times and locations of mutants before $t=0$ also
well--approximated by a 
Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Poisson Mapping Theorem, a Poisson process with rate we define to be $\lambda_0$.
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}
Note that this is closely related to the equilibrium frequency of a
allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}

\paragraph{Geographic spread of alleles} 
Once an allele has become locally established it can begin to spread across space. 
We assume that the allele, once established, 
quickly settles down to spread spatially as a traveling wave of constant speed. 
The behavior of this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. 
Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
See \citet{ralph2010parallel} for a fuller review of these travelling waves.
Note that the speed of the wave will vary
with details of the space that individuals migrate across
\citep[e.g.\ see ][ for comparisons to migration on discrete grids]{Slatkin-speed:76,SlatkinCharlesworth:78}.

@@

@@

Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutatoin at different locations within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Three G6PD deficiency alleles that also confer resistance to malaria 
are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years  \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$
\citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
This is in agreement with the selection coefficients calculated by
\citet{Ruwende-g6pd} on the basis of the present day levels of
resistance to malaria due to the $A-$ allele. \\

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. 
The pie chart shows the frequency of G6pd-deficiency alleles. 
The size of the pie chart indicates the number of G6pd-deficient individuals sampled.
Countries with endemic malaria are colored yellow. 
Figure taken from \citet{Howes-g6pd-variants}
\url{http://www.malariajournal.com/content/12/1/418}. 
} \label{fig-G6PD-map}
\end{center}
\end{figure}

@@

Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
As it spreads, it will cover a distance $L$ in time $L/v$.
The number of other mutations appearing in the circle it has covered up until this time is Poisson with mean
$\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
(and $\lambda_0 2 L + \lambda \pi L^2 /v$ in one dimension).
% $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
\begin{equation}
    \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
\end{equation}
then $\chi$ gives the distance spread unobstructed by the descendants of a new mutant
before it is expected that one other successful mutation would have arisen in the area covered so far.
The explicit formula for $\chi$ is cumbersome; here we omit it.
(In one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
we can rewrite this as
\begin{equation} \label{eqn:defines_chi}
%    \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^2 + \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^3 
%        = \frac{ - \log(1-s_d)^3 }{ 4 s_b^2 \sigma^2 \pi \rho \mu } .
   \frac{\sqrt{2s_b} }{-\log(1-s_d) } \chi^2 + \frac{1}{\sigma} \chi^3 = \frac{1}{\sqrt{2s_b} \rho\mu\pi}
\end{equation}
From this we see that $\chi$ decreases with $\rho$ and $\mu$.
Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
\begin{equation} \label{eqn:chi_standing}
\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma).
\end{equation}
On the other hand, if the mutant allele is highly deleterious before $t=0$,
then the characteristic length is close to the value from \citet{ralphcoop2010}:
\begin{equation} \label{eqn:chi_new}
\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d)).
\end{equation}
Theses two end points help build our intuition for the interaction of
parameters in shaping the geographic scale of convergent evolution.
By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
 and occur within the first $\chi/v$ generations.
 Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
 we expect to find one mutational origin.

%XXX could do calculations as in "panmixia" here, e.g.\ if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
%$f = 1-\gamma f^{3/2}$, where
%$\gamma = (4 \pi^{5/2} / \sqrt{2}) \mu^2 s_b^{3/2} \rho^2 / \sigma \log(1/(1-s_d))$.
%Then if $\gamma \ll 1$, $f \approx 1$, while if $\gamma \gg 1$, then $f \approx 0$.

%%%GRAHAM: I commented this out. We return to do proportion standing
%%%below, so perhaps nicer to restrict that to one place. 
% By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
% and occur within the first $\chi/v$ generations.
% Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
% we expect to find one mutational origin;
% we denote the probability that this mutation is from standing variation by $\pi_0$.
% Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
% and it is an property of competing Poisson processes that
% \begin{align} \label{eqn:pizero}
%     \pi_0 = \lambda_0 \pi \chi^2 .
% \end{align}
% (In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)

In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb} we show the range of characteristic lengths as a function 
of various parameters chosen to match the evolution of malaria
resistance at G6pd. 
These curves match our intuition that population densities result in
smaller characteristic lengths (as would higher mutation rates). 
Unsurprisingly, allowing standing variation increases the mutational
input and so decreases the characteristic length below that predicted
by only new mutations (i.e. eqn. \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
As the prior deleterious effect of the allele ($s_d$), the
characteristic length increases untill it reaches that predicted by
new mutation alone. Higher dispersal distances lead to larger
characteristic lengths, as alleles spread geographically more rapidly blocking others
from arising. A larger selective advantage acts in two conflicting
ways: aiding the rapid geographic spread of established alleles, and also help more independent copies
escape drift and become established. The effect on helping establish
alleles wins out in this as increasing the selective $s_b$ decreases
the characteristic length. This effect strongest when only standing
variation contributes ( $s_b^{-1/2}$, eqn \eqref{eqn:chi_standing}), 
as in that case the speed of spread does not matter, the dependence
when only new mutations contribute is much weaker ( $s_b^{-1/6}$, eqn \eqref{eqn:chi_new})
Overall, the range of characteristic lengths observed are reasonably
consistent with the average diameter of a G6PD variant in Eurasia of
$800$, especially for the lower population density.

@@

The quantity that is perhaps easiest to compute is the mean time until adaptation.
If fix ourselves at some geographic location, and let $\tau \ge 0$ be the time at which the point is reached by the advantageous mutations.
Then, as can be seen from Figure~\ref{fig:cartoon},
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\begin{equation}
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\end{equation}
and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

In Figure \ref{G6PD_chartimes} we show the mean time until adaptation
for various values of the parameters chosen to match the case of
adaptation at G6PD. Increasing $\sigma$ and decreasing $s_d$ lower 
the time to adaptation, as alleles spread geographically more quickly
and are present as standing variation more often
respectively. Increasing $s_b$ strongly decreases the time to
adaptation, as it both causes more alleles to escape drift and to
rapidly spread. Given that the G6PD alleles likely spread over a few
thousand years, i.e. less than a few hundred generations. This
time-scale seems quite plausible given Figure \ref{G6PD_chartimes},
except perhaps for the lowest dispersal distances.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
\caption{ %
{\bf Mean adaptation times}, as a function of selective disadvantage
and , compared to the corresponding quantity without standing variation, and to the quantity only considering standing variants.
} \label{G6PD_chartimes}
\end{center}
\end{figure}

@@

\subsection{The contribution of standing variation.}
We can use our framework to address how much adaptation arises from
standing mutation. We have defined $\lambda_0$ to be the mean density
of standing variants that are present and escape loss when the
environment shifts.
We will define $\nu_+$ to be the mean density of newly arising alleles
that spread --having arisen in an area free of other adaptive alleles.
Since the probability that a mutant arising at $(x,t)$ is lucky enough to be born in a location not already occupied by mutants
is $\P\{ \tau > t \}$,
we can see  $\nu_+ = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
$\nu_+ = \lambda \E[\tau] $.
Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
\begin{equation}
    \nu_+ = (\mbox{mean density of new patches}) = \frac{1}{1-\log(1-s_d) \E[\tau]} .
\end{equation}
Therefore, the mean proportion of patches that come from standing variation
is 
\begin{equation} \label{prop_patches_standing}
\lambda_0 / (\lambda_0 + \nu_+),
\end{equation}
this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
There are $\lambda_0 + \nu_+$ patches per unit area, so
the typical patch (i.e.\ with distribution given by the Palm measure) occupies area $1/(\lambda_0 + \nu_+)$.

We can also find the mean proportion of space covered by standing variants.
If by time $t$ a geographic location has not yet been reached by the mutation,
the probability it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$, 
and the probability it is reached by a new variant is $\lambda \pi v^2 t^2 dt$.
Therefore, as is standard for competing exponentials,
the probability a given location is reached first by a standing variant,
and therefore the mean proportion of space covered  by standing variants,
is
\begin{equation} \label{prop_space_standing}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt .
\end{equation}
To evaluate this integral, again see Appendix \ref{apx:integrals}.
%\marginnote{put in this value in terms of Gaussian CDF?}

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
\caption{ %
{\bf Mean proportion of patches arising from standing variation.} The
parameters given are for the G6PD example.
The left and right panels show the expected proportion of patches that
arise from standing variation and the expected proportion of
geographic space that is covered by adaptation from standing variation
(eqn. \eqref{prop_patches_standing} and \eqref{prop_space_standing} respectively).
} \label{G6PD_standing_var_proportion}
\end{center}
\end{figure}

%{\tt Simulations:} 
%compare mean density against $s_d$ at different values of $\sigma$; 
%omit variability if needed for readability. 
%Compare to panmictic results.

\subsubsection{Multiple variant types}
Another problem that we can seek to address with this work is
the extent to which pleiotropy biases 
adaptation towards the repeated use of particular subset of loci
(i.e.\ convergence at genetic level). 
While many alleles may confer the benefical phenotype, 
not all will contribute to adaptation if they have negative pleiotropic consequences.
There are at least two ways that negative pleiotropy can contribute to high rates of convergence 
when adapting if a single change is sufficient for adaptation. 
First, negative pleiotropic effects can reduce 
the overall beneficial selection coefficient of an allele in the new environment,
making them unlikely to become established and slow to spread 
(and in the worst case making them deleterious). 
This effect has been studied by a number of authors \citep{REFS}. 
A second contribution is that alleles that have
less negative pleiotropy are more likely to be present as standing variation 
before the environmental shift.

Here we focus primarily on the second effect. 
Let's imagine for the moment that all classes of beneficial allele 
have the same beneficial selection coefficient. Or simply beneficial
selection coefficients are similar enough that our selective exclusion
approximation holds over the time-scales over which we examine the process.
Each class of mutations could have its own mutation rate $\mu_j$ and
have a disadvantage $s_{d,j}$ prior to the switch. 
As they have the same beneficial selection coefficient after the switch, 
all of the waves travel outward at a rate $v$. 
Then the density of type $k$ standing variants per unit area 
and the input rate of \textit{de novo} variants per unit area per generation are, respectively,
\begin{align}
  \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } \qquad \text{and} \qquad    \lambda_{j} = \mu_j \rho s_b .
\end{align}
Using these rates it follows that, at the time when every location has been reached by an adaptive allele,
the proportion of the species range covered by alleles of type $k$ is
\begin{equation} \label{eqn-prop-space-allele-k}
    p_k = \int_0^\infty  \left( 2 \lambda_{0,k} \pi v^2 t + \lambda_k \pi v^2 t^3 \right)  
    \; \exp \left( - \sum_j ( \lambda_{0,j} \pi v^2 t^2 + \lambda_j \pi v^2 t^3 / 3 ) \right) dt .
\end{equation}
If we only allow standing variation this collapses to 
\begin{equation}  \label{eqn-prop-space-allele-k-standing-room-only}
    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) } ,
\end{equation}
while if we only allow new variation, 
i.e.\ if all variants are highly deleterious before the environment switches, 
then $p_k =\mu_j / (\sum_k  \mu_k)$.
